\section{Introduction}%
\label{sec:introduction}

\subsection{Motivation}%
\label{sub:motivation}

Model theory concerns itself with the connection between first order theories, that is
sets of axioms possibly using universal and existential quantifiers, and the models of these
theories. Some early examples of significant results in model theory can be found in the
compactness and Löwenheim–Skolem theorems, all of which focused on the existence of models for a
theory: the compactness theorem saying that a model exists for a theory if models
exist for each finite fragment of the theory; and the Löwenheim–Skolem theorems saying
that if a theory has an infinite model, then there exist models of any cardinality
greater than or equal to the number of symbols in the language.

As a consequence of the Löwenheim–Skolem theorems, we know that there exists no theory
with a unique infinite model up to isomorphism. This raises the question of what can
be said about the number of models a theory has for each cardinality.

\begin{defn}
  For a complete theory $\theory$, we write $I(\theory, \kappa)$ to mean the number of
  models, up to isomorphism, of $\theory$ with cardinality $\kappa$. We call
  $I(\theory, -)$ the spectrum of $\theory$.
\end{defn}

One of the first answers to this question came in the form of Morley's categoricity
theorem, which said that for a complete countable theory $\theory$, if $I(\theory,\kappa) = 1$ for
some uncountable $\kappa$, then $I(\theory,\lambda) = 1$ for all uncountable $\lambda$. \cite{10.2307/1994188}
Work by Shelah in the 1970s then extended this result to uncountable theories
\cite{Sh:31}, beginning the study of classification theory as a discipline of model theory.

Important notions from stability theory came in the form of stable and unstable theories.
Stable theories are well-behaved enough to limit their number of models, while unstable theories
always have the maximum number of models, that is if $\theory$ is unstable then
for any $\kappa > |\theory|$ we have $I(\theory,\kappa) = 2^\kappa$. \cite{Sh:a}
Most naturally occuring
theories in mathematics tend not to be stable though, for example the full theory of any linear
order or the theory of real closed fields. Hence it is only natural to study generalisations of
stable theories, which will not be as well-behaved but will hopefully allow us to learn about
wider classes of theories. One such generalisation comes in the form of the non-independence
property, which informally says that there is no formula which can pick out every subset of
an infinite subset of a model.

Classification theory is not the only active discipline of model theory however. Another
area with a lot of interesting questions comes from the study of homogeneous structures,
where any isomorphism between finite substructures can be extended to an automorphism of
the homogeneous structure. In 1953, Fraïssé showed how to construct such structures
by gluing classes of finite structures together \cite{fraisse1953certaines},
from which their study followed.

Hence, in the spirit of these disciplines, we will consider the first-order theory of
interval algebras, which were originally introduced by Allen in 1983 to argue about time
qualitatively, and try to find their place in the universe of model
theory, by considering their different models and how well-behaved they are.

\subsection{Report Structure}%
\label{sub:report_structure}

The rest of the report will be structured as follows. In
\cref{sec:background} we summarise the results from Allen's original paper
on interval algebras \cite{allen83}, focusing on the choices that lead to
the 13 relations and the algorithm to infer missing relations from a given
interval network. We also introduce the necessary notions from model theory
such as homogeneous models, and stable/NIP theories. We try to keep this
introduction grounded by working through some examples where possible.
Readers comfortable with these areas may safely skip them in favour of
\cref{sec:axiomatisation} and onwards.

In \cref{sec:axiomatisation} we tackle the question of axiomatisation of
interval algebras. We propose some axioms based on the understanding from
\cref{sec:axiomatisation} and show that these are satisfiable by
constructing models from linear orders. To further justify this
axiomatisation, we will then construct linear orders from interval
algebras.

In \cref{sec:adjunction} we explore further the intervals and points
constructions from \cref{sec:axiomatisation} and show they can be extended to two adjoint
functors between the relevant categories of models. We also give characterisations of
the interval algebras and linear orders for which the unit and counit are isomorphisms.
These characterisations will be expressible as first order sentences, a fact we will use
when studying the classification theory of interval algebras.

Finally, in \cref{sec:model_theory} we study the model theory of interval algebra. This
begins by considering the the class of finite interval algebras and computing its
Fraïssé limit using the machinery developed in \cref{sec:adjunction}. On the topic of
stability, we show that the stable interval algebras are exactly the finite interval
algebras. Following this there is a study of the NIP in interval algebras, for which
we find a big class of NIP algebras as well as an example of an interval algebra with the
IP.

\subsection{Ethical Considerations}

Model theory, being a subfield of mathematical logic, is an area certainly very theoretical and
firmly lodged inside ``pure mathematics''. As such, the work done in this project has been
entirely theoretical and mainly for aesthetic purposes. While Allen's interval algebra is a very
concrete construction with a lot of possible real-world applications, none of these applications
are expected to benefit from our findings.

And so the main ethical consideration to be had is one familiar to all theoretical projects, that
of correctness. Because if our work is both wrong and has no applications, then it will really
have been for nothing. Hence a lot of care has been taken to ensure the whole content of this report
is correct and that all the arguments hold up to scrutiny. Any omitted proofs have still been
checked on paper and so, short of formalising all of this work in a theorem prover, we can be quite
sure of its correctness.
